---
title: "Exam Project"
author: "Filip **********"
date: "30-Nov-21"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    code_folding: show
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: yes
mainfont: null
sansfont: null
mathfont: null
monofont: null
fontsize: null
email: *****@student.uu.se
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts = list(width.cutoff = 60))
```


# Data management 
This section must include: data import, variable assignment, dataset reorganisation (merge + long format)

```{r chunk-1}
## Library imports
library(dplyr) # Import dplyr for pipelines
library(ggplot2) # Import ggplot2 for boxplots
library(ncappc) # Import for alternative, automated AUC and k calculation
library(GGally) # Import for pairplot/correlations


## Functions
# Calculate total body water dependent on sex
Calculate_TBW <- function(Sex, Age, Height, Weight) {
  if (Sex == "M") {
    TBW <- 2.447 - 0.09156 * Age + 0.1074 * Height + 0.3362 * Weight
  } else
    TBW <- -2.097 + 0.1067 * Height + 0.2466 * Weight
  return(TBW)
}

# Numerical summary
Summarize <- function(column) {
  print("Mean:")
  print(mean(column))
  print("Median:")
  print(median(column))
  print("Standard deviation:")
  print(sd(column))
  print("Range:")
  print(range(column))
}

# Correlation pair plots
Plot_Correlations <- function(var1, var2, var3, Headline) {
  plot <- ggpairs(data_PK_SNP[c(var1,var2,var3)], aes(colour = data_PK_SNP$Sex, alpha = 0.5))
  plot <- plot + theme_minimal() + ggtitle(Headline)
  print(plot)
}

# Boxplots for t1/2 and AUC for SNPS
Plot_t12_AUC_Boxplots <- function(HalfLife, AreaUnderCurve, Headline) {
  col_names <- names(data_PK_SNP) # Get column names
  plot_list <- list() # Create empty list for appending plots
  min_AUC <- min(AreaUnderCurve) # Find AUC minimum for combined plot
  max_AUC <- max(AreaUnderCurve) # Find AUC maximum for combined plot
  
  # Plot for AUC by ks
  for (i in c(21:25)) { # Plot for different mutations
    plot_list[[i - 20]] <- eval(substitute(
      ggplot(data_PK_SNP, aes(group = data_PK_SNP[, c(i)], color = data_PK_SNP[, c(i)], fill = data_PK_SNP[, c(i)]))
      + geom_boxplot(aes(y = HalfLife, AreaUnderCurve), alpha = 0.6)
      + facet_grid(cols = vars(data_PK_SNP[,c(i)])) 
      + labs(x = "AUC [mg / L * h]", y = "t 1/2 [h]", title = col_names[i])
      + xlim(c(min_AUC, max_AUC)) # Same x axis for all mutations for better comparison
      + theme_minimal() + theme(legend.position = "None") # Set theme and remove legend
      + scale_color_manual(values = c("#ff800e", "#006ba4", "#006ba4"))
      + scale_fill_manual(values = c("#ff800e", "#006ba4", "#006ba4")),
      list(i = i)
    ))
  }
  
  # Arrange box plots in grid
  plot <- ggpubr::ggarrange(plotlist = plot_list, ncol = 3, nrow = 2)
  ggpubr::annotate_figure(plot, top = Headline)
  
}

# Plot t_1/2 against TBW with linear regression
Plot_t12_TBW <- function(HalfLife, Headline) {
  plot <- ggplot(data_PK_SNP, aes(x = TBW, y = HalfLife, color = Sex)) # Coloring by sex
  plot <- plot + geom_point(alpha = 0.7) + geom_smooth(method = "lm") # Create scatter plot + add linear regression
  plot <- plot + labs(x = "Total body water [L]", y = "t 1/2 [h]") + theme_minimal() # Add labeling + theme
  plot <- plot + ggtitle(Headline)
  print(plot)
}

# Create path + file name variable
dir <- "" # Assume you're in the working directory or change this
file_name_PK <- "BPI889_PK_17.csv"
file_name_SNP <- "BPI889_SNP_17.txt"

path_PK <- paste(dir, file_name_PK, sep = '')
path_SNP <- paste(dir, file_name_SNP, sep = '')

## Import files into data.frame and assign variables
data_PK <- read.table(path_PK, sep = ",", header = T, na.strings = ".") # Blood concentrations
# separator = "," since the data is comma separated
# header = T since a header is included which will be our column names
# na.strings = "." since missing values are indicated by dots (".")

data_SNP <- read.table(path_SNP, header = T, skipNul = T) # Polymorphism data
# header = T since a header is included which will be our column names
# skipNul = T since tab separator ("\t") doesn't work to get the data in the correct columns

# Rename patient ID column
ind_X <- which(colnames(data_PK) == "X") # Find index of column with name "X"
names(data_PK)[ind_X] <- "ID" # Reassign column name to "ID" for data_PK
data_SNP <- tibble::rownames_to_column(data_SNP, "ID") # Make rownames first column for data_SNP

# Combine data_PK and data_SNP
data_PK_SNP <- merge(data_PK, data_SNP, by = "ID")

# Get time points from column names
time_pts <- as.numeric(colnames(data_PK_SNP) %>% { gsub("Time.", "", .) } %>% { gsub(".h", "", .) })
names(data_PK)[2:16] <- time_pts[2:16] # Rename columns

## Convert merged data from wide to long format while removing NAs
# Full data in long format
data_PK_SNP_long <- tidyr::gather(data_PK_SNP, value = VALUE, key = KEY,-ID, na.rm = T)
```


# Variable calculations
This section must include: calculation of body size measurement, categorization of body size measurement, PK variable calculation

```{r chunk-2}
# Calculate total body water + find cmax 
ind_start <- which(colnames(data_PK) == 0.15) # Find index of column with start concentration at t = 0.15 h
ind_end <- which(colnames(data_PK) == 24) # Find index of column with end concentration at t = 24 h

for (i in 1:nrow(data_PK_SNP)) { # Loop TBW calculation through all patients for TBW, cmax calculation
  data_PK_SNP$TBW[i] <- Calculate_TBW(data_PK_SNP$Sex[i], data_PK_SNP$Age..yrs.[i], data_PK_SNP$Height..cm.[i], data_PK_SNP$Weight..kg.[i]) # Calculate and append TBW
  data_PK_SNP$c_max[i] <- max(data_PK_SNP[i, ind_start:ind_end], na.rm = T) # Find and append cmax
}

# Split into TBW < 40 L and TBW > 40 L
data_PK_SNP$TBW_grouped <- data_PK_SNP$TBW # copy data to replace by 0s and 1s for categories
data_PK_SNP$TBW_grouped[data_PK_SNP$TBW_grouped < 40] <- 0
data_PK_SNP$TBW_grouped[data_PK_SNP$TBW_grouped > 40] <- 1
data_PK_SNP$TBW_grouped <- factor(data_PK_SNP$TBW_grouped, labels = c("TBW < 40 L", "TBW > 40 L"), levels = c(0,1)) # Factorize and relabel

## Calculate t1/2, AUC, k
na_24h <- which(is.na(data_PK_SNP$Time.24.h)) # Rows/patients where t = 24 h is missing -> Extrapolating necessary
na_015h <- which(is.na(data_PK_SNP$Time.0.15.h)) # Rows/patients where t = 24 h is missing -> Extrapolating necessary

# Extrapolate 24 h data
for (i in na_24h) {
  x <- time_pts[14:15]
  y <- as.numeric(data_PK_SNP[i, 14:15])
  data <- data.frame(y = y, x = x) # Combine data
  model <- lm(y ~ x, data) # Create linear model
  pred_24h <- predict(model, newdata =  data.frame(x = time_pts[16]))
  if (pred_24h < 0) {
  data_PK_SNP$Time.24.h[i]  <- 0 # Set to 0 because negative concentrations aren't possible
  } else {
  data_PK_SNP$Time.24.h[i]  <- pred_24h  # Set to extrapolated concentration
  }
}

# Extrapolate 0.15 h data
for (i in na_015h) {
  x <- time_pts[3:4]
  y <- as.numeric(data_PK_SNP[i, 3:4])
  data <- data.frame(y = y, x = x) # Combine data
  model <- lm(y ~ x, data) # Create linear model
  pred_015h <- predict(model, newdata =  data.frame(x = time_pts[2]))
  if (pred_015h < 0) {
    data_PK_SNP$Time.0.15.h[i]  <- 0 # Set to 0 because negative concentrations aren't possible
  } else {
    data_PK_SNP$Time.0.15.h[i]  <- pred_015h  # Set to extrapolated concentration
  }
}

# Interpolate other missing values
missing_NAs <- list() # Create empty list to fill
for (i in 1:ncol(data_PK_SNP)) {
  missing_NAs[[i]] <- which(is.na(data_PK_SNP[i])) # Create list with indices of all missing values
}

for (time in 2:16) { # Loop through all time points
    if (length(missing_NAs[[time]]) == 0) { # Skip where no values are missing
      time <- time + 1 
    } else {  
      for (i in unlist(missing_NAs[time])) { # Loop through missing values
      y <- data_PK_SNP[i, (time - 1):(time + 1)] # Values before and after the missing value
      x <- time_pts[(time - 1):(time + 1)] # Timepoints before and after the missing value
      pol <- approx(x, y, n = 3) # Calculate interpoalted value
      data_PK_SNP[i,time] <- pol$y[2] # Replace NA with interpolated number
    }
  }
}

# Calculate ks 
for (patient in 1:nrow(data_PK_SNP)) { # Loop through all patients
  time <- 16 # Begin search for non-zero concentrations at latest time point (column 16)
  while (time > 2) { # Loop through all time points 
    if (data_PK_SNP[patient, time] & data_PK_SNP[patient, time - 1] != 0) { # Until two neigbohring non-zero concentrations are found
      t <- time_pts[time] - time_pts[time - 1] # Calculate time difference
      data_PK_SNP$k[patient] <- 
        abs((log(data_PK_SNP[patient, time]) - log(data_PK_SNP[patient, time - 1])) / t) # Calculate "k" and make positive
      break
    } else {
      time <- time - 1 # Go to next time point
    }
  }
}

# Calculate half-life
data_PK_SNP$t_12 <- log(2) / data_PK_SNP$k 

# Calculate AUC from Trs and ks
Trs <- list()
for (i in 3:16) { # Loop over all concentrations/time points
  Trs[i - 2] <- (data_PK_SNP[i] + data_PK_SNP[i - 1]) / 2 * (time_pts[i] - time_pts[i - 1])
}

Trs <- do.call(cbind, Trs) # Create data.frame from list
data_PK_SNP$AUC <- rowSums(Trs)/data_PK_SNP$k # Calculate AUC from sum of TRs and k
 
## Calculate AUC, t_1/2 and k by package ncappc
# Time and concentration in long format
data_PK[2:16] <- data_PK_SNP[2:16] # Copy with inter-/extrapolated values
data_conc_long <- tidyr::gather(data_PK[1:16], value = Concentration, key = Time, -ID, na.rm = T)
data_conc_long$Time <- as.numeric(data_conc_long$Time) # Make time points numeric
names(data_conc_long) <- c("ID", "TIME", "DV") # Rename columns for ncappc

ncappc(obsFile = data_conc_long, onlyNCA = T, extrapolate = T, printOut = F, evid = FALSE, noPlot = T) # let ncappc do the work
data_PK_SNP$k_by_package <- ncaOutput$Lambda_z # Get k
data_PK_SNP$t_12_by_package <- ncaOutput$HL_Lambda_z # Get t1/2
data_PK_SNP$AUC_by_package <- ncaOutput$AUClast # Get AUC
```


# Data exploration
This section must include: numerical summary of PK variables, graphical assessment of 1) PK profiles, 2) PK variable correlations, 3) PK variable-SNP correlations, 4) PK variable-body size measurement correlation with linear regression


```{r chunk-3}
# Numerical summary
Summarize(data_PK_SNP$c_max)
Summarize(data_PK_SNP$t_12)
Summarize(data_PK_SNP$AUC)
Summarize(data_PK_SNP$t_12_by_package)
Summarize(data_PK_SNP$AUC_by_package)

## Graphical assessment
# PK profiles
min_conc <- min(data_PK_SNP[2:16]) # Find minimum concentration for combined plot

for (i in c(21:25)) {
  data_PK_SNP[,i] <- factor(data_PK_SNP[,i], labels = c("Wildtype", "Heterogeneous", "Homogenous")) # Factorize and rename
}

# Ggplot Line Plot
ggplot(data = data_conc_long, aes(x = TIME, y = DV, group = ID)) +
  geom_line() + theme_minimal() + ylab("Concentration [mg / L]") + xlab("Time [h")

# Create correlation pairplots
Plot_Correlations(27, 30, 31, "Values from my script") # Plot for script values
Plot_Correlations(27, 33, 34, "Values from ncappc") # Plot for ncappc values

# # PK variable-SNP correlations (Boxplots)
Plot_t12_AUC_Boxplots(data_PK_SNP$t_12, data_PK_SNP$AUC, "Values from my script")
Plot_t12_AUC_Boxplots(data_PK_SNP$t_12_by_package, data_PK_SNP$AUC_by_package, "Values from ncappc")

# PK variable-body size measurement correlation with linear regression
Plot_t12_TBW(data_PK_SNP$t_12, "Values from my script")
Plot_t12_TBW(data_PK_SNP$t_12_by_package, "Values from ncappc")
```

# Statistical testing
This section must include: ANOVA of PK variables for SNPs, t-test of PK variable for body size measurement groups

```{r chunk-4}
result_anova_AUC <- aov(AUC ~ T134A + A443G + G769C + G955C + A990C , data = data_PK_SNP)
summary(result_anova_AUC) # for values from script

result_anova_AUC <- aov(AUC_by_package ~ T134A + A443G + G769C + G955C + A990C , data = data_PK_SNP)
summary(result_anova_AUC) # for values from ncappc

result_anova_cmax <- aov(c_max ~ T134A + A443G + G769C + G955C + A990C , data = data_PK_SNP)
summary(result_anova_cmax) # for cmax

# T-Test (t1/2 for TBW > 40 L and TBW < 40 L)
t.test(data_PK_SNP$t_12 ~ data_PK_SNP$TBW_grouped) # for values from script
t.test(data_PK_SNP$t_12_by_package ~ data_PK_SNP$TBW_grouped) # for values from ncappc
```
